Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-08-10 Ben_Manuscripts/hdphmm/response_to_reviewers.tex

    As stated by the reviewer, there are two goals of our work: to obtain long timescale MSDs and
    to understand mechanisms leading to diffusive behavior. TEKMC may be a suitable alternative 
    approach to obtaining our first goal, but it is not clear that it could give the same level
    of mechanistic insight that the HDP-AR-HMM offers. Although the work by Hanson et al.
    (J. Phys Chem. B, 2012, 116, 95-103) is able to provide physical insight into the effect of
    nanoparticles on diffusion of gaseous penetrants in a polystyrene polymer matrix, it does so
    by testing hypotheses motivated by observations of the diffusion constant (predicted by TEKMC)
    in response to different temperatures and nanoparticle concentrations. This is actually pretty
    in line with the type of analysis we performed in our initial LLC transport work (J. Phys Chem. B, 2019, 123, 6314-6330)
    where we used the mean squared displacement and the qualitative characteristics of the solute
    time series in order to hypothesize transport mechanisms. In this work we "re-discover" these
    mechanisms with minimal human intervention and characterize them in detail using the 
    HDP-AR-HMM while \textit{simultaneously} providing a method to estimate diffusion constants.

%MRS: additional thoughts:

%1. Dynamic Bond Percolation theory (both the original version and the
%chemically specific one) an and TEKMC are based on spatial clustering
%and coarse-graining. Our approach is based on clustering and
%coarse-graining the dynamics.  Can be more powerful, doesn't assume
%(or minimally assumes - in some cases we assume pore vs. nono-pore)
%which configurations are important, which matters when there are a
%diversity of different types of bound states, and diversity of
%dynamics (in the pore, associated with the head groups, etc).

%2. Dynamical behavior is more complex with our solutes than ion
%interactions because of the richer range of interactions involved
%(ionic binding PLUS hydrophobic binding, i.e. more types of bound
%states; it's not just binding and hopping, and thus something

%3. Configurational behavior is more complex with the nanostructures,
%because one cannot simply divide the structure into cubes
%isotropically.  It may be easier to use our approach with more
%diverse configurations, because we don't need to worry about
%partitioning correctly.  This does push some of the difficulty into
%choosing the coordinate system for the dynamics, but we have a lot
%more freedom about choosing coordinate systems for the model than one
%would using a uniform (or even nonuniform) partitioning of the
%system. Yes, there are difficult choices to make but the model has
%the freedom to make these choices.

%4. One reason our simulations are longer because the timescales of
%events observed generally are longer, and the number of solutes we
%can put into the simulation is smaller (since the total pore volume
%is smaller than it would be in an amorphous polymer).

%5. We could probably construct our model it with a larger number of
%more shorter simulations as well. Rare events require long timescale
%sisulations.  However, the equilibration times for these simulations
%(since we build from geometry, not from amorphous melts) are such
%that we probably get more efficient use of simulations by running
%fewer, longer simulations.

%6. So it's not clear that this method is better than DBP or TEKMC
% (especially since the results are a big ambiguous); one could define
% bound states, and look at hops betwen then.  But it's definitely
% DIFFERENT, it's not really the same thing at all, and we should push
% back that it is similar.  We can argue that our approach is more
% suited to our systems than the two configuration-based approaches -
% one COULD try to use those approaches and they could be made to
% work, but it's not a trivial thing to do, and we wanted to try
% something where the data informed our model more completely and we
% wouldn't have to try those modifications.

@@

    We do agree that there are certain basic simulation details that we have neglected to 
    mention. Therefore we have made the following edits to section 2.1 of the methods:

    We added the highlighted text to the first sentence of the section:
    \begin{quote}
      We studied transport of solutes in the H\textsubscript{II} phase using \hl{our
      previously developed} atomistic molecular model of four pores in a 
      \hl{$9.4\times9.4\times7.6$ nm} monoclinic unit cell with 10\% water by weight, 
      \hl{totaling 62,000 atoms}.    
    \end{quote}

    We have also added the following text describing the model:
    \begin{quote}
    We modeled water molecules using the TIP3P water model,~\cite{jorgensen_comparison_1983}
    We used the Antechamber module shipped with AmberTools18 to parameterize monomers and 
    solutes using the Generalized Amber Force Field (GAFF).~\cite{wang_development_2004} We 
    assigned charges using the AM1-BCC model as implemented in Antechamber.~\cite{jakalian_fast_2000,jakalian_fast_2002}
    GAFF is very commonly used in simulations with diverse organic chemical functionality, 
    and was developed for consistency within the AMBER ecosystem, which uses TIP3P as the
    water model. We chose these standard force fields and generalizable parameterization 
    approaches so that a range of solutes and monomers could be self-consistently and easily 
    studied. The main goal of this work is to develop stochastic models which can reproduce
    solute behavior on MD time scales, independent of the choice of force field. In future 
    work, it may be beneficial to employ specialized force fields, including polarizable 
    force fields or other higher-order effects in order to improve the accuracy of structure 
    parameters and selectivity predictions.
    \end{quote}

    Finally, we have added a figure with a picture of the system which we are studying:
    \begin{figure}
    \centering
    \includegraphics[width=\textwidth]{membrane_side_profile.pdf}
    \caption{When combined with 10 wt\% water, the LLC monomer studied in this work forms the 
    H\textsubscript{II} phase which is characterized by hexagonally packed columns with 
    hydrophilic pores. In the figure, white and red water molecules mark the membrane pores.
    The monomer fills the rest of the space. Note that water molecules are present in the 
    tails, but they are hidden for the sake of clarity. }\label{fig:membrane_side_profile}
    \end{figure}

@@

    Author reply: We thank the reviewer for posing this question as this a constraint to which 
    we gave thought but neglected to emphasize in the main text. 
    %BJC: he's not saying that solutes should stay in the aqueous pore region right? We talk pretty
    % extensively about solute behavior in the tails so his steric argument is confusing me.

    %MRS: It is a little confusing. First, we never use the term "pore
    %radius", since the structure of the pore makes it very hard to
    %define a single pore radius.  The term "pore region" is used once
    %(about line 995), that that should probably be "pore center".
    %BJC2: changed pore region to pore center
    %Second, it's not clear what is meant when they say it's not clear
    %how they are constrained when they must be by steric
    %interactions.  Since steric interactions do NOT keep them in the
    %pore region (at least for the smallest solutes).  Figure 2 is a selection of a trajectory.

    %    First, we would like to clarify
%    that our simulations suggest that small solutes can partition out of the pores and into
%    the dense tail region. This is not a particularly favorable move as evidenced by our radial
%    distribution functions, but it does happen and could conceivably lead to rare jumps of solutes
%    between pores (although we have not observed any).

    %MRS: how about this?  
    %BJC2: seems good. made some minor edits
    We are grateful for the opportunity to clarify some of the details of the simulation 
    geometries. We avoid defining a ``pore radius'' because the radial distribution of 
    head groups and even tail groups is spread over several nanometers and we have
    observed clear evidence that solutes can similarly partition radially throughout 
    the pores.~\cite{coscia_understanding_2019,coscia_chemically_2019} 
%    This partitioning occurs to the extent that smaller molecules like water and methanol
%    can in fact move between pores in rare instances.
    %BJC2: I don't think I've actually published anything stating or proving the above. 
    %I think it's particularly hard to get by the crosslinks. But I do believe it could happen.
    %rephrased with that in mind.
    Therefore, it may be possible that, in rare instances, smaller molecules like water and
    methanol may move between pores. However the rarity of these events and the fact that 
    they occur orthogonal to the transport dimension means that they do not significantly 
    affect transport, and can be interpreted mathematically as simply going deep into the 
    tail region before returning to the same pore.

@@

    Because of this, our stochastic simulations are constrained to remain close to a single 
    pore by the trajectory generation procedure. Each state is parameterized by a mean in 
    the radial direction computed based on the solute's distance from the \textit{nearest 
    pore center}, making it impossible for the solute to get further than approximately half
    the pore-to-pore distance away (the exact maximum distance is an an angle-dependent 
    quantity due to the hexagonally packed pore configuration).
    %MRS: though I'm not QUITE sure that's right, since the hexagonal packing means that the transition between pores
    %occurs at different distances depending on the angle. 
    %BJC: Yeah, I know what you mean. I added the bit in parentheses above to acknowledge that.
    Each time a state change occurs, the mean position of the solute in the $r$ dimension 
    jumps to the appropriate mean. However, in the axial direction, we do not parameterize 
    the mean since it should be unbounded in that dimension. The $z$ mean of each segment 
    is set based on the position immediately before the state change occurred.

    We have strengthened Section 2.3 of the methods by adding a clearer discussion of the above.

    Original text:
    \begin{quote}
    We obtained the mean vectors, $\mathbf{c}$, of the clustered states by averaging each 
    value of the $\mathbf{c}$ vectors assigned to the same cluster. Note that we only care 
    about the $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$ because solute trajectories are not 
    bound in the $z$ direction.
    \end{quote}

    Modified text:

    \begin{quote}
    We obtained the mean vectors, $\mathbf{c}$, of the clustered states by averaging each 
    value of the $\mathbf{c}$ vectors assigned to the same cluster. Note that we only care 
    about the $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$. We do not parameterize the mean
    $z$ position of particles because solute trajectories should not be bound to discrete
    points in the axial direction. However, we do intentionally limit the radial location 
    of particles in our stochastic trajectories to discrete values of $r$, effectively 
    constraining them to stay within a single pore. While it's possible that, in rare cases,
    solutes might move between adjacent pores, it would have no effect on axial transport
    since this motion is orthogonal to the pore axis and because all pores are structurally
    similar. This treatment assumes that a move between pores is equivalent to a solute 
    which travels deep into the monomer tail region before returning to the same pore. 
    \end{quote}

@@

    The generalization of this procedure to three dimensions will be under active investigation 
    in the near future. As mentioned, we are interested in applying this approach to bicontinuous
    cubic phases like the gyroid. Based on our current understanding, we would model solute dynamics
    in three dimensions and set the mean vector, $\mathbf{c}$ equal to $\mathbf{0}$. Of course, this
    may result in solutes exploring a three dimensional space that does not share the geometric 
    features of the studied phase. Our hypothesis is that we can add one or more additional 
    dimensions to the time series passed to the HDP-AR-HMM that includes a distance from the pore centers. 
    We could define the pore center using a spline or a surface and transitions would likely
    favor moves towards states that exist closer to the pore interiors. Alternatively, it may prove
    necessary to constrain solute positions to the means in each dimension. However, these extensions
    are beyond the scope of the current paper.

    We have added the following text to the end of Section 2.3 in order to make it clear that
    the trajectory generation procedure will require more thought for different geometries:
    \begin{quote}
    The parameterization of $\mathbf{c}$, $A$ and $\Sigma$ as presented above has been designed 
    specifically for the cylindrical pore geometry at hand. Our group is actively investigating
    appropriate ways to parameterize more complicated geometries such as bicontinuous cubic phase
    LLC membranes. 
    \end{quote}

@@

    %r1_comment7
    \begin{quote}
    For the purposes of our analysis, \sout{we assume ergodicity of the 24 solute trajectories. That is,}
    we assume our MD simulations sample all possible states with the correct frequency.
    \end{quote}

    \item \textit{There are no units on Figure 1 y-axis, presumably these are nanometer like the 
    other figures?}

    Author reply: We thank the reviewer for catching our oversight. We double-checked the 
    other plots in the paper and found the same mistake on the y-axes of Figure 5. Units of nanometers
    have been added to the y-axes of both plots.

    %r1_comment8
    \item \textit{Figure captions need to be improved. Wording such as “we plotted...”, “we 
    calculated...”, “We also analyzed...” should be omitted in figure captions.}

    %BJC2: okay, modified as suggested
    Author reply: We agree with the reviewer that the type of wording described should not be
    included in captions as their primary purpose should be to explain the conclusions drawn
    by the figure rather the process of generating the figure. However, we believe that the 
    only real offender among our captions was Figure 8. Therefore we have modified the language
    of that figure:

    Original caption:
    \begin{quote}
    (a) We analyzed hydrogen bonds donated by methanol to the highlighted oxygen
    atoms in three regions of the monomer: the carboxylate groups (blue), the ether linkages
    (green) and the tails (red). (b) For each predominant state (states 1--5 described in 
    Figure~\ref{M-fig:prevalence}), we plotted the fraction of the total time in that state 
    spent hydrogen bonding to various regions of the monomer, color coded to match (a).
    (c) For each type of hydrogen bond in each predominant state, we calculated the 95th percentile
    of hydrogen bond lifetimes. (d) We also analyzed the frequency and lifetimes of solute 
    association with sodium ions. Methanol associates with sodium less than 10\% of the total 
    time it occupies any of the predominant states.
    \end{quote}

    Modified caption:
    \begin{quote}
    (a) Hydrogen bonds can be donated by methanol to the highlighted oxygen atoms in three 
    regions of the monomer: the carboxylate groups (blue), the ether linkages (green) and the
    tails (red). (b) For each predominant state (states 1--5 described in Figure~\ref{M-fig:prevalence}),
    we show the fraction of the total time in that state spent hydrogen bonding to various regions
    of the monomer, color coded to match (a). (c) For each type of hydrogen bond in each predominant
    state, we show the 95th percentile of hydrogen bond lifetimes. (d) We also show the 
    frequency and lifetimes of solute association with sodium ions. Methanol associates with sodium
    less than 10\% of the total time it occupies any of the predominant states.    
    \end{quote}

@@

    %r2_comment1
    \item \textit{This paper presents an approach to study transport mechanisms of solutes within
    membranes by applying the sticky hierarchical Dirichlet process autoregressive hidden Markov model
    (HDP-AR-HMM) to molecular dynamics simulation trajectories spanning long times (5 microseconds). 
    Using this method, the authors reveal different dynamic states associated with the transport of 
    solute across a membrane, and interpret them on the basis of solute-membrane interactions. The 
    paper is clear, well written, and relevant to the Journal of Physical Chemistry. I only have a 
    couple of suggestions and minor corrections:}

    Author reply: We thank the reviewer for their kind assessment of our work.

    %r2_comment2
    \item \textit{A motivation for the study is providing a computational tool to guide experimentalists
    in designing membrane materials. However, given the length of the molecular dynamics simulations 
    required, it is presumably impractical to carry a large-scale screening in this way. It would be 
    useful if the authors were more explicit about the role of this approach in guiding experiments, 
    and what would be its advantages and limitations compared to a purely experimental search.}

    % BJC2: shifted initial focus towards mechnistic understanding. Still kept part about speeding up simulations
    % as I think it's worth discussing ways to make this an actual high throughput technique. 
    Author reply: We thank the reviewer for pushing us to further evaluate the practicality of our
    method as a tool for experimentalists. The length of our simulations in this paper are indeed 
    quite long. 
    %The dynamics of the solutes are somewhat slow and we wanted to ensure we gathered
    %enough data for a sufficient parameterization. 
    However, the length of the simulations primarily influences the reliability of the long
    timescale predictions of our model. We believe one of our method's greatest strengths is its 
    ability to elucidate transport mechanisms by automatically detecting shifts in dynamical
    behavior, which can be achieved with much shorter simulations. 
    In our previous work, we showed that one can use shorter simulations to uncover the same 
    mechanisms which we described in our current work.~\cite{coscia_chemically_2019} However, the
    HDP-AR-HMM makes it much simpler to identify these mechanisms. Our approach allows one to 
    quickly identify behavior characteristic of solutes in terms of chemical functionality, 
    allowing them to eliminate sets of potential LLC monomers on the basis of this more general
    mechanistic understanding.

    There is also value in considering ways to make high throughput simulation of this type of 
    system attainable. Putting aside brute force high throughput simulations that could be enabled 
    by the ever-increasing availability of HPC resources, it would be to our benefit to use the
    system in this work as a benchmark to explore much less expensive models. We propose some ideas
    to this end in the text we have added to the end of Section 2.1, where our simulations are described:
    \begin{quote}
    In consideration of our ultimate goal, which is to use high-throughput simulations to 
    facilitate our understanding of LLC membrane design for solute-specific separations, the size 
    of our system and length of the accompanying molecular simulations appears to be somewhat 
    impractical. Dependent on one's goals, it may not be necessary to run such long simulations in order 
    to extract valuable mechanistic insight. We originally uncovered the mechanisms discussed
    in this study using trajectories a fraction of the length here.\cite{coscia_chemically_2019} A 
    strong mechanistic understanding of solute transport with respect to the chemical features of a 
    given LLC monomer may be enough for experimentalists to eliminate entire subsets of potential
    design changes. However, it may still be valuable to consider how one might make these simulations
    amenable to high-throughput studies. We would like to emphasize that our large, fully atomistic 
    models will be extremely valuable as a benchmark system used to explore less expensive models. 
    For example, we could explore systems built with fewer LLC monomers by making the membrane pores
    shorter. We could also build systems with higher concentrations of solutes in order to gather more 
    dynamical data on shorter timescales. Using what we have learned about the radial distribution
    functions of these solutes in this and our previous work~\cite{coscia_chemically_2019}, it may be possible to 
    learn how to create initial configurations with solutes initially distributed in a way that
    is close to their equilibrium distribution, allowing solutes to thoroughly explore the 
    accessible membrane structural space more quickly and reducing the amount of unequilibrated
    data that must be discarded. Alternatively, we could use grand canonical Monte Carlo simulations
    to load solutes into the membrane~\cite{snurr_prediction_1993}. Finally, it may be possible 
    to speed up our simulations by reducing the number of atoms using united-atom representations
    of the liquid crystal monomers~\cite{horta_gromos-compatible_2016} or by using hydrogen mass 
    repartitioning to allow us to increase the MD time step.~\cite{hopkins_long-time-step_2015}
    \end{quote}

@@

    %r2_comment3
    \item \textit{Have the authors attempted to compare their results directly to experimental data? 
    It would be useful to clarify whether the predicted selectivities are expected to be quantitatively
    accurate or just reflect trends.}

### 2020-08-10 Ben_Manuscripts/hdphmm/response_to_reviewers.tex

    %BJC: This is copied straight from the reviewer response for the previous paper.
    %BJC2: do you think I should reword this at all to make it different?
    Author reply: The reviewer makes a valid point. This work started in parallel with experimental
    work which promised solute flux and selectivities that we could compare directly to our
	simulations. However, synthesizing an aligned hexagonal phase membrane has been a practical
	challenge for our experimental collaborators and much of their focus has shifted to 
	membranes formed by liquid crystals that self-assemble into bicontinuous cubic architectures
	which are much simpler to synthesize. We have also started to shift our focus towards models 
	of the bicontinuous cubic phase, which do have much more experimental data. However, our 
	research into generating physically accurate molecular models of that phase lags our work on
	time series modeling. In future work, we will be able to apply this technique to these 
	bicontinuous systems for which we have experimental data.

### 2020-08-10 Ben_Manuscripts/hdphmm/Draft.tex
Edits, new figure, new references

  %BJC: reworked this section for better clarity as it confused Chris
  \textit{Obtaining Parameters of Clustered States}: We obtained the 
  mean vectors, $\mathbf{c}$, of the clustered states by averaging each 
  value of the $\mathbf{c}$ vectors assigned to the same cluster. Note 
  that we only care about the $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$ 
  because solute trajectories are not bound in the $z$ direction.

@@

%  In Section~\ref{S-section:ihmm_procedure} of the Supporting
%  Information, we provide a step-by-step illustration of the following process.

>>>>

  %BJC: r1_comment5
  \textit{Obtaining Parameters of Clustered States}: We obtained the mean vectors,
  $\mathbf{c}$, of the clustered states by averaging each value of the $\mathbf{c}$
  vectors assigned to the same cluster. Note that we only care about the 
  $r=\sqrt{x^2+y^2}$ portion of $\mathbf{c}$. We do not parameterize the mean
  $z$ position of particles because solute trajectories should not be bound to discrete
  points in the axial direction. However, we do intentionally limit the radial location 
  of particles in our stochastic trajectories to discrete values of $r$, effectively 
  constraining them to stay within a single pore. While it's possible that, in rare cases,
  solutes might move between adjacent pores, it would have no effect on axial transport
  since this motion is orthogonal to the pore axis and because all pores are structurally
  similar. This treatment assumes that a move between pores is equivalent to a solute 
  which travels deep into the monomer tail region before returning to the same pore.

@@

  %r1_comment5
  The parameterization of $\mathbf{c}$, $A$ and $\Sigma$ as presented above has been designed 
  specifically for the cylindrical pore geometry at hand. Our group is actively investigating
  appropriate ways to parameterize more complicated geometries such as bicontinuous cubic phase
  LLC membranes.
